---
title: "Diversity Index for Maryland Cities"
date: "Nov 8, 2025"
author: "Your Name"
---

**Your Name Here**
Eden Binder

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Overview

In this exercise, we will calculate the **USA TODAY Diversity Index** for Maryland cities with a population of at least 10,000 using 2020 Census race and ethnicity data. The Diversity Index reflects the probability that two randomly selected individuals from a city are of different races or ethnicities.

A score of 0 indicates no diversity (everyone is the same race/ethnicity), while a score closer to 100 indicates higher diversity. The **most diverse city** has the highest index, reflecting a balanced mix of several racial and ethnic groups. The **least diverse** city likely has one overwhelmingly dominant racial/ethnic group.

## Task 1: Load and Describe the Data (3 points)

Run the following code to load the data, then examine it and describe what it contains. Based on the raw numbers alone, which cities stand out as diverse to you?


```{r load-data}
md_cities <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/usat_diversity_index.csv")

# Preview structure
View(md_cities)
```

**ANSWER HERE**
This data conains information on the population numbers of cities within Maryland based on racial identity. The data shows the total population, along with how many people identify as two or more races, Black, white, Hispanic, Native American, Asian, Native Hawaian and Pacific Islander, other and not Hispanic. Based on raw numbers alone cities such as Baltimore, Bowie, Silver Spring, Wheaton, Montgomery Village, and Columbia stand out as diverse to me.

## Task 2: Calculate Race Percentages

```{r race-percentages}
md_cities <- md_cities |>
  mutate(
    White_pct = White / TotalPop,
    Black_pct = Black / TotalPop,
    Native_pct = Native / TotalPop,
    Asian_pct = Asian / TotalPop,
    NHPI_pct = NHPI / TotalPop
  )
```

## Task 3: Square the Race Percentages

We need to square the race percentages in order to calculate the probability that any two people chosen at random from a given census area are of different races or are Hispanic or not.

```{r square-race}
md_cities <- md_cities |>
  mutate(
    White_pct_sq = White_pct^2,
    Black_pct_sq = Black_pct^2,
    Native_pct_sq = Native_pct^2,
    Asian_pct_sq = Asian_pct^2,
    NHPI_pct_sq = NHPI_pct^2
  )
```


## Task 4: Calculate Race Probabilities (3 points)

The race probability is the sum the squared probabilities for each of the races. It represents the probability that two people randomly chosen from the same city will be of the same race. Which city has the highest race probability, and why do you think that is?


```{r race-prob}
md_cities <- md_cities |>
  mutate(
    Race_Prob = White_pct_sq + Black_pct_sq + Native_pct_sq + Asian_pct_sq + NHPI_pct_sq
  ) |> 
  arrange(desc(Race_Prob))

View(md_cities)
```

**ANSWER HERE**
The city of Summerfield has probability that two people randomly chosen will be of the same race. I think this is because Summerfield's population is over 93% Black, which is the highest percentage of a single race amongst all cities in Maryland, so it makes sense that the city would subsquently have the highest race probability.
## Task 5: Calculate Ethnicity Percentages and Their Squares (3 points)

Hispanic origin is not a race but an ethnicity or nationality – and a separate Census question. The probability someone is Hispanic or not Hispanic must be calculated separately from race, and Hisp_Prob is the probability that two randomly selected people will be Hispanic or not. That's an important distinction, because unlike race probability, this measures the likelihood of a binary question: that both random people will be Hispanic. What do you notice about the cities with the *lowest* Hisp_Prob scores?


```{r ethnicity}
md_cities <- md_cities |>
  mutate(
    HispTotal = HISPANIC + NotHispanic,
    Hisp_pct = HISPANIC / HispTotal,
    NotHisp_pct = NotHispanic / HispTotal,
    Hisp_pct_sq = Hisp_pct^2,
    NotHisp_pct_sq = NotHisp_pct^2,
    Hisp_Prob = Hisp_pct_sq + NotHisp_pct_sq
  ) |> 
  arrange(Hisp_Prob)

View(md_cities)
```

**ANSWER HERE**
The cities with the lowest Hispanic probability all have other recial and ethnic identities with signifigantproportion, which likely lowers the Hispanic probability even with some of the cities having a population that has a high proportion people who identify as Hispanic.
## Task 6: Multiply Race and Ethnicity Probabilities

Here we multiply the two probabilities together to get the probability that any two random people have the SAME race and ethnicity.

```{r combined-prob}
md_cities <- md_cities |>
  mutate(
    RaceEth_Prob = Race_Prob * Hisp_Prob
  )
```


## Task 7: Calculate Diversity Index (3 points)

The final step is to calculate the probability that two people randomly chosen WILL NOT BE OF THE SAME RACE OR ETHNICITY. In other words, the highest index scores will be cities where those two random people will be from different backgrounds. Put simply, more diverse. Which cities are the top 3, and where are they located? Look at the bottom of the list - which city is the least diverse, and where is it?


```{r diversity-index}
md_cities <- md_cities |>
  mutate(
    Diversity_Index = (1 - RaceEth_Prob) * 100
  ) |> 
  arrange(desc(Diversity_Index))

View(md_cities)
```

**ANSWER HERE**
Langley Park of Prince George's County, Wheaton of Montgomery County and Adelphi of Prince George's County have the top 3 diversity index scores. Lake Shore of Anne Arundel County is the least diverse.
### Pitch a story (8 points)

Given the data that you've produced here, write a headline and lede of the most interesting story from this exercise, and add below that details on how you would report it out: who you'd talk to and the questions you'd ask. When you describe the Diversity Index values, do so in a similar way to how USA Today did; the raw values themselves won't mean much to readers.

**ANSWER HERE**
Headline: Prince George's County reaches both highs and lows for diversity within Maryland
Lede: According to the USA Today diversity index that calculates the probability that two randomly chosen people within one city will not be of the same race or ethnicity, Prince George's County holds both two of the top three most diverse cities and two of the top three least diverse cities within the state.

I would mainly go about reporting this story by seeing what I could do in terms of talking to people within all four of these communities within Prince George's to see if I could get their thoughts in terms of how they feel about the diversity (or lack of thereof) that is present in their communities, as well as what sort of difference, if any it makes in their lives. The type of people I would want to talk to would definitely be just regular community members, but probably also teachers, religious leaders and politicans who may be able to give their thoughts on the substance behind these numbers due to their roles as leaders or those who frequently interact with fellow community members. I think my questions would be more directed towards if these community members felt the data was accurately reflected, what do they think their communities can do to increase or maintain current diversity levels, and are they satisfied with the current amenities and quality of life that their communities offer, or would they perhaps like to see improvement, and if so in what areas. 
