
---
title: "Multiple Regression Analysis in R"
date: 8-24-2025
---

# Introduction

Multiple regression analysis is a statistical method used to understand the relationship between one dependent variable and two or more independent variables. This analysis is particularly useful in journalism for exploring complex relationships, such as how different factors collectively impact outcomes like election results, economic indicators, or sports performances. By understanding these relationships, journalists can provide deeper insights and more accurate predictions.

```{r setup}
library(tidyverse)
install.packages('corrplot')
library(corrplot)
```

## Data Import

The first step in our analysis is to load the data. Here we read the data from a CSV file, which contains various statistics for NFL teams from the 2019 season. This dataset will be used to build our model.

```{r}
teams <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/main/nfl_2019.csv")
```

## Model Building

Next, we construct a linear model to analyze how different factors like `Yards Gained`, `Takeaways`, and `Points Scored` influence the number of wins. This model helps us quantify the relationships and assess which factors are significant predictors of success.

```{r}
model <- lm(Wins ~ `Yards Gained` + Takeaways + `Points Scored`, data = teams)
summary(model)
```

### Residuals

The residuals of a model are the differences between observed values and the values predicted by the model. For your model:

*   **Min and Max** values indicate the range of residuals. A maximum of 4.1361 and a minimum of -4.3780 suggest that there might be some substantial errors in predictions for some cases, but generally, the errors are relatively evenly distributed around zero.
*   **Median** close to zero (-0.0604) indicates that the model is generally centered around the actual data points, but still, some deviations exist.

### Coefficients

The coefficients tell us the estimated effect of each independent variable on the dependent variable:

*   **Yards Gained**: The coefficient is -0.001329. This suggests a very small negative impact of yards gained on wins, but this effect is not statistically significant (p = 0.34332), implying that as yards gained increase, the number of wins slightly decreases, which is counterintuitive and likely indicates no real effect or that the model isn't capturing the complexities.
*   **Takeaways**: The coefficient is 0.150593, suggesting that each additional takeaway is associated with an increase in wins by about 0.151, which is near the threshold of significance (p = 0.07928). This suggests a potentially meaningful, though not conventionally statistically significant, positive effect on wins.
*   **Points Scored**: The coefficient is 0.038732, which is significant (p = 0.00665). This indicates that each additional point scored is associated with an increase in wins by about 0.039, suggesting a clear and meaningful impact.

### Model Fit

*   **Residual Standard Error**: The RSE of 2.072 tells us that the typical prediction by the model deviates from the actual number of wins by about 2.072 wins.
*   **Multiple R-squared**: 0.6232 indicates that about 62.32% of the variability in wins is explained by the model, which is relatively high, suggesting a good fit.
*   **Adjusted R-squared**: 0.5828 adjusts the R-squared for the number of predictors and indicates that about 58.28% of the variance is explained by the model after accounting for the number of variables used.
*   **F-statistic**: The F-statistic of 15.44 with a very low p-value (4.058e-06) suggests the model as a whole is statistically significant.

Is this a useful model, based on what you've seen? Why?

I would say that the model above is a useful one. Looking at the multiple R-squared value it tells you that 62% of the time, a majority, of the variation in NFL team wins is explained by the model, which is a relatively high amount and indicates a good fit. Additionally, although yards gained and takeaways are proven to not be statistically significant predictors when it comes to wins according to their respective p-values, points scored are shown to be a statistically signifigant and meaningful predictor of team wins. The model's Residual standard error or RSE of 2.072 showing that the model deviates from each team's win total by just over two wins, which is a very good margin given that each team at this point played 16 games a season.

## Correlation Analysis

After building the model, we calculate the correlation matrix for our variables of interest. This matrix helps visualize the strength and direction of the relationships between each pair of variables.

```{r}
selected_vars <- teams |>  select(`Yards Gained`, Takeaways, `Points Scored`, Wins)
cor(selected_vars)
```

## Visualization of Correlation Matrix

Finally, we use a visualization tool to plot the correlation matrix. This graphical representation makes it easier to interpret the correlations at a glance, showing which variables are most closely connected.

```{r}
corr_matrix <- cor(selected_vars)
corrplot(corr_matrix, method = "circle")
```

## Results Interpretation

The results from our correlation analysis show the strength of the relationships between the number of wins and factors like yards gained, takeaways, and points scored. High positive correlations indicate strong relationships where increases in one variable are associated with increases in another. Conversely, negative correlations suggest an inverse relationship. By understanding these dynamics, journalists can better explain the factors that contribute to a team's success or failure in a season.

Which of these three factors - Yards Gained, Takeaways and Points Scored - have the most impact on wins? Are there any that do not seem impactful at all?

Points scored has the most impact on wins out of the three above. Yards gained does not seem to impact wins at all, while takeaways are not that statistically impact on wins, its p-value of  0.07928 puts it very close to the signifigance or impact threshold of 0.05. The only factor that does not seem impactful at all is the yards gained factor.

## Using the Model

Without using the words "multiple regression model", describe how you could use this model to cover a team or game.

To cover a team or game and use this model without mentioning the words you can as a journalist you could explain how factors such as takeaways or yards gained are not proving to affect the specific game results or win totals of the team that you are covering, while points scored has proven to be an accurate indicator of your teams' success or lack thereof.
