---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 0.5?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:**
Our group's mean was not close to half at all. Other colored M&Ms dominated the amount of red and black ones. Different groups are getting different results due to each group having a different number of M&Ms. When we combine all groups such as within one bag og M&Ms there is less varience due to there being a more equal of each of the six M&M colors than there would be in one handful. In terms of real life this can show how random sampling with too small of a sample size can produced highly varied/extreme results that may not reflect the reality.
------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:**
Only about 22.8% of the UMD Community responded to the survey. Student response rate was only at 20%, while staff response rate was at 36%, making it so students were likely underrepresented. The survey also had 4 percent less Hispanic representation than the campus population, 2 percent less African American representation, 5 percent more people that had a missing or unknown ethnicity, and 6 percent of people who identify as multiracial. In order to deal with these differences, the survey says they used qualitative responses instead of quantitative responses that were analyzed using content analysis to extract common themes and thematic patterning across respondents.I would say that one question for me is the survey's decision to group together some clashing racial or ethnic identities such as in the missing/unknown category and multiracial category.
------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(98)
people <- tibble(
  age = sample(18:80, 1000, replace = TRUE)
)

head(people)

```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}
sample_100 <-people |> 
  sample_n(100)

sample_50 <-people |> 
  sample_n(50)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}
sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))
sample_50 |> summarize(mean = mean(age), median = median(age), sd = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
people |> summarize(mean = mean(age), median = median(age), sd = sd(age))
```


### Task 6: Compare the sample and population statistics
**A**
```{r}
The sample and population statistics are almost the same when it comes to the mean, median, and sd. The sample mean is slightly smaller than the population. The sample median as well as the sd is slightly larger than the population.

```

Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

The sample mean and sd is slightly smaller than the population. When running the sampling and statistics calculations with a smaller sample size, the mean grows even smaller than the larger sample and population, while the sd is larger than both as well.

**A**:

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
